Old or Heavy?\\Decaying Gracefully with Age/Weight Shapes.

### 2019-05-28 papers/CADE19/paper.tex
CADE final

\documentclass{llncs}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{amssymb}
\usepackage{amsmath}

@@

\begin{document}
\maketitle
\begin{abstract}
Modern saturation theorem provers are based on the given-clause algorithm, which iteratively selects new clauses to process. This clause selection has a large impact on the performance of proof search and has been the subject of much folklore.  
The standard approach is to alternate between selecting the \emph{oldest} clause and the \emph{lightest} clause with a fixed, but configurable \emph{age/weight ratio} (AWR).
We show that an optimal fixed value of this ratio can produce proofs significantly more quickly on a given problem, and further that varying AWR during proof search can improve upon a fixed ratio.
Based on these observations we develop several new modes for the Vampire prover  which vary AWR according to a ``shape'' during proof search.
The modes solve a number of new problems in the TPTP benchmark.
\end{abstract}

@@

Currently, the most successful theorem provers (such as Vampire~\cite{vampire}, E~\cite{E}, or SPASS~\cite{Spass}) for first-order logic are saturation-based, utilising the well-known \emph{given-clause algorithm}.
Simply, this algorithm \emph{saturates} a set of clauses by iteratively selecting a clause and performing all non-redundant inferences with it until all clauses have been selected or the empty clause (witnessing inconsistency) has been found.
Clearly, the order in which clauses are selected is key to the performance of the algorithm.
Over the past few decades a certain amount of folklore has built up around the best methods for clause selection and recent work by \sandm{} \cite{clause-selection-heuristics} systematically studied these.
Our work extends this study with new results and also introduces the concept of a \emph{variable} clause selection strategy (one that changes over time) which we instantiate with two simple patterns (or \emph{shapes}) that prove to be pragmatically useful.

We restrict our attention to clause selection strategies that alternate between selecting clauses based on age (i.e. in a first-in-first-out manner) and weight (i.e. those with the fewest symbols first).
It was confirmed by \sandm{} that alternating these two heuristics outperforms either by itself.
We refer to the ratio with which we use either heuristic as the \emph{age/weight ratio} (AWR) as this is the terminology employed by the Vampire theorem prover, which we use as a vehicle for our study.

After covering relevant background material in Section~\ref{sec:background} the remainder of the paper makes two main contributions.
Firstly, in Section~\ref{sec:awr:study} we experimentally confirm the folklore that (i) the choice of age/weight ratio often has a significant effect on the performance of proof search, and (ii) there is no `best' age/weight ratio: indeed, a large range of pragmatically useful ratios exists.
Secondly, we demonstrate (Section~\ref{sec:varying:study}) that varying the age-weight ratio over time can achieve better performance than a fixed ratio and use this to motivate the addition of so-called \emph{age/weight shapes} for varying the ratio over time.
Experiments (Section~\ref{sec:experiments}) with these new options implemented in the Vampire theorem prover show a significant improvement, proving many new problems unsolvable by any current configuration of Vampire.

@@

\paragraph{First-Order Logic and Weight.} Our setting is the standard first-order predicate logic with equality. A formal definition of this logic is not required for this paper but an important notion is that of the \emph{weight} of a clause. In first-order logic,  terms are built from function symbols and variables, literals are built from terms, and clauses are disjunctions of literals. The weight of a term is the number of function and variable symbols occurring in it and this notion is lifted to clauses.

@@

\paragraph{The Given Clause Algorithm and AWR Clause Selection.}
To achieve saturation, the \emph{given clause algorithm} organises the set of clauses into two sets: the \Active\ clauses are those that have been active in inferences, whereas the \Passive\ clauses are those that have not. Typically, a further \Unprocessed\ set is required in order to manage the clauses produced during a single iteration of the loop. 
Realisations of the given clause algorithm generally differ in how they organise simplifications. There are two main approaches (both implemented by Vampire, originally found in the eponymous theorem provers Otter~\cite{otter} and \textsc{Discount}~\cite{discount}): the Otter loop uses both \Active\ and \Passive\ for simplifications, whereas the Discount loop only uses \Active.

The algorithm is centred around the \emph{clause selection} process. As previously mentioned, there are two main heuristics for this:
\begin{itemize}
	\item \emph{By Age (or First-in/First-out)} clause selection prefers the oldest clause (produced earlier in proof search), simulating a \emph{breadth-first} search of the clause space. In Vampire the age of a clause is the number of inferences performed to produce it (e.g. input clauses have age 0).
	\item \emph{By Weight (or symbol-counting)} clause selection prefers the youngest clause (with the fewest symbols). The intuition behind this approach is that the sought empty clause has zero symbols and lighter clauses are (in some sense) closer to this. Furthermore, lighter clauses are more general in terms of subsumption and tend to have fewer children, making them less explosive in terms of proof search.
\end{itemize}
\sandm~show that alternating these heuristics is beneficial and we consider this setting here. In Vampire this alternation is achieved by an age/weight ratio (AWR) implemented by a simple \emph{balancing} algorithm. The balance is initialised to 0 and used as follows: a positive balance means that a clause should be selected by age, whereas a negative balance means that a clause should be selected by weight; given a ratio of $a:w$ the balance is incremented by $a$ when selecting by age and decremented by $w$ when selecting by weight.
Figure~\ref{fig:discount} gives the Discount algorithm along with balance-based AWR clause selection. The lines relevant to clause selection are marked with \Mark.

\begin{figure}[t]
\begin{tabbing}
~~~~~~\=~~~~\=~~~~\=~~~\=~~~~\=~~~~\=~~~~\=~~~~\=~~~~\=\kill
\>\INPUT: $\Init$: set of clauses\semicol, $a:w$ age-weight ratio \\
\>\VAR\ $\Active$, $\Passive$, $\Unprocessed$: set of clauses\semicol \\
\>\VAR\ $\Given$, $\New$: clause\semicol\\
\>$\Active\ \assign\ \emptyset$\semicol \\
\>$\Unprocessed\ \assign\ \Init$\semicol \\
\Mark
\>$\Balance\ \assign\ 0$\semicol\\
\>\LOOP\\ 
\>\>\WHILE\ $\Unprocessed \neq \emptyset$ \\
\>\>\>$\New \assign \Pop(\Unprocessed)$\semicol \\ 
\>\>\>\IF\ $\New = \emptyclause$ \THEN\ \RETURN\ \textit{unsatisfiable}\semicol \\
\>\>\>\IF\ $\Retained(\New)$ \THEN \` (* retention test *) \\ %\inc
\>\>\>\>simplify $\New$ by clauses in $\Active \cup \Passive$ \semicol 
                        \` (* forward simplification *) \\
\>\>\>\>\IF\ $\New = \emptyclause$ \THEN\ \RETURN\ \textit{unsatisfiable}\semicol \\
\>\>\>\>\IF\ $\Retained(\New)$ \THEN
                        \` (* another retention test *) \\
\>\>\>\>\>simplify $\Active$ and  $\Passive$ using $\New$ \semicol
                        \` (* backward simplification *) \\
\>\>\>\>\>move the simplified clauses to $\Unprocessed$\semicol \\
\>\>\>\>\>add $\New$ to $\Passive$\\ %\dec 
\>\>\IF\ $\Passive = \emptyset$ \THEN\
          \RETURN\ \textit{satisfiable} or \textit{unknown}\\
\Mark
\>\>\IF\ $\Balance > 0$ \THEN\\
\Mark
\>\>\>\>$\Given\ \assign\ $ lightest clause in \Passive\semicol\\
\Mark
\>\>\>\>$\Balance \assign\ \Balance - w$\semicol\\
\Mark
\>\>\ELSE\\
\Mark
\>\>\>\>$\Given\ \assign\ $ oldest clause in \Passive\semicol\\
\Mark
\>\>\>\>$\Balance \assign\ \Balance + a$\semicol\\
\>\>move $\Given$ from $\Passive$ to $\Active$\semicol \\
\>\>$\Unprocessed \assign \Infer(\Given,\Active)$\semicol \` (* generating inferences *)\\
\end{tabbing}

@@

@@

We note here that Vampire's portfolio mode includes an additional option relevant to clause selection: the \texttt{--nongoal\_weight\_coefficient} option specifies a multiplier to apply to the weight of non-goal clauses, thus preferring clauses in or derived from the goal in clause selection. Use of this heuristic is orthogonal to the age/weight ratio and we do not investigate its impact any further here.

%%%%%%
\paragraph{Related Work.}
We briefly review the clause selection approaches taken by other solvers. 
Otter 3.3~\cite{otter3} selects \emph{either} by age, by weight or manually.
Prover9~\cite{prover9} allows a configurable age/weight ratio.
E~\cite{E} allows the user to specify an arbitrary number of priority queues and a weighted round-robin scheme that determines how many clauses are picked from each queue.
The default is to use a combination of age and weight selection, although there is also a complex strategy developed by a genetic algorithm~\cite{genetic-heuristics}. 
SPASS~\cite{Spass} uses symbol-counting based clause selection. 
iProver~\cite{Iprover} follows in E in having a number of configurable queues but relies mainly on age and weight heuristics in those queues. 
The general idea in this paper of a \emph{varying age/weight ratio over time} is applicable to any ratio-based clause selection strategy, and our specific results apply to those that take a ratio between age and weight.

\section{Optimising Age/Weight Ratios}
\label{sec:awr:study}

We set out to experimentally confirm two assumptions from folklore:
\begin{enumerate}
	\item The choice of age/weight ratio often has a significant effect on the performance of proof search.
	\item There is in general no single best age/weight ratio for a given set of problems.
\end{enumerate}
These were supported by the work of \sandm{} but here we focus on these points and explore them more deeply.

\subsection{Logarithmic AWR}

Optimising AWR values is more easily achieved if they have a continuous scale.
AWR values are mathematically \(\mathbb{Q}^{+}\), the positive rational numbers, but in practice are more easily seen logarithmically: 1:16 is the same distance in terms of the effect on proof search from 1:8 as it is from 1:32.
Therefore, we introduce the \emph{logarithmic} AWR \(L\), defined in terms of age \(A\) and weight \(W\) as
\[
	L = \log_2{\left(\frac{A}{W}\right)}
\]
With this definition, AWR can now be seen continuously.
As \(L\) tends to positive infinity, Vampire selects only by weight, whereas if \(L\) tends to negative infinity Vampire selects only by age.
\(L = 0\) represents the middle ground of a 1:1 age/weight ratio. Note that we have not updated the balancing algorithm used by Vampire to make use of this value (this still requires two numbers) but use the number in experiments when discussing AWR.

\begin{figure}[t]
	\centering
	\includegraphics[width=0.8\textwidth]{example-optimal-awr}
	\caption{
The number of activations reported by Vampire after successful 1-second runs on a TPTP problem.
In between the peaks on either side, the function of \(L\) is discontinuous with large peaks and troughs, but follows an approximate trend and settles toward the global optimum.
This is typical for the observed set of problems.
}
	\label{fig:example-optimal-awr}
\end{figure}

\subsection{Experiments}
As an initial illustrative example of how varying the AWR effects the number of clauses required to be processed before a proof is found consider Figure \ref{fig:example-optimal-awr}. This demonstrates the effect that varying AWR can have: a smaller number of activations means that fewer clauses were processed, which in general means that a proof was found faster\footnote{
It should be noted that if a small number of clauses are extremely expensive to process it may be slower than a larger number of less-expensive clauses, but in general this is a good heuristic measure for prover performance.
It also avoids reproducibility issues involved with using system timing approaches.
}.
On this problem, a good AWR value is over 400\% better by this metric than the worst AWR value.

\begin{table}[t]
	\caption{
Relative performance gain, showing the ratio in activations between the best AWR setting for a given problem and another base setting.
We compare  1:1 (Vampire's default), 1:5 (the best-behaved from \sandm), and the worst setting for the problem.
Where the problem is not solved at all by the base setting, it is ignored.
}
	\centering
	\begin{tabular}{c | c | c c}
		& Maximum Gain & Mean Gain & (Standard Deviation)\\
		\hline
		1:1 & 133.56 & 1.26 & 1.63 \\
		1:5 & 133.67 & 1.44 & 1.70 \\
		(worst) & 222.01 & 3.95 & 7.60 \\
	\end{tabular}
	\label{tab:awr-improvement}
\end{table}

This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not currently support (e.g. higher-order problems). We ran Vampire for 1 second in default mode with the \texttt{discount} saturation algorithm\footnote{The default LRS~\cite{LRS} saturation algorithm can be non-deterministic so we avoid it for such experiments.} using a sensible set of AWR values (see Table \ref{tab:no-best-awr}) --- these are the values used in Vampire's portfolio mode. These tend to favour weight-first over age-first as this has been experimentally shown to be preferable. We then filter out problems not solved by any of these or those solved trivially (e.g in preprocessing). The whole set yielded similar AWR data for 7,947 problems.

The first result is that choosing a good AWR value for a problem is well-rewarded.
Table~\ref{tab:awr-improvement} summarises the impact that choosing the best AWR can have.
Compared to the default we can perform, on average, 1.26 times fewer activations, which is modest but (as Table \ref{tab:no-best-awr}) shows we also lose just under 10\% of problems by choosing the default.
It is more relevant to note that there are cases where we can do \emph{much} better by selecting a different AWR value.
So choosing a better AWR value \emph{can} go from no solution to a solution and \emph{can} do so faster, but not necessarily.
In the worst case (choosing the pessimal AWR value) we can perform almost 4 times as many activations.

The second result is that there is no ``best'' AWR across this full set of problems. Table \ref{tab:no-best-awr} shows, for each AWR value, the \% of problems solved, the number solved uniquely, the maximum and mean drop in performance. The drop in performance is how many times more activations were required for a proof in the worst/average case compared to the best. No AWR value solves all problems, with the best being 1:5. 
A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
\sandm~found that 1:5 had a similar property, but this might be explained by differences in prover and test environment. It is interesting to note that the extreme AWR values solve fewer problems overall but solve the most uniquely. This is typical in saturation-based proof search: approaches that do not perform well in general may perform well in specific cases where the general approach does not.

@@

\begin{table}[t]
	\caption{
Per-AWR value results on 1 second runs over 7,947 TPTP problems.
	}
	\centering
	\begin{tabular}{l | c | c | c | c  c }
		AWR & \% Solved & Uniques & Maximum Drop & Mean Drop & (Standard Deviation)\\
\hline
8:1 & 85.25 & 16 & 150.67 & 1.37 & 1.98\\
5:1 & 86.10 & 1 & 122.22 & 1.33 & 1.64\\
4:1 & 86.93 & 1 & 101.44 & 1.32 & 1.42\\
3:1 & 87.63 & 2 & 105.00 & 1.29 & 1.41\\
2:1 & 88.62 & 3 & 112.67 & 1.27 & 1.45\\
3:2 & 89.83 & 2 & 119.89 & 1.27 & 1.51\\
5:4 & 89.98 & 4 & 125.00 & 1.26 & 1.55\\
1   & 90.56 & 4 & 133.56 & 1.26 & 1.63\\
2:3 & 91.20 & 9 & 147.67 & 1.28 & 1.79\\
2   & 91.68 & 0 & 162.67 & 1.31 & 1.97\\
3   & 91.81 & 5 & 190.56 & 1.37 & 2.30\\
4   & 91.85 & 3 & \textbf{17.41} & 1.38 & 0.67\\
5   & 92.00 & 2 & 133.67 & 1.44 & 1.70\\
6   & 91.57 & 1 & 106.44 & 1.47 & 1.46\\
7   & 91.49 & 1 & 104.89 & 1.49 & 1.44\\
8   & 91.09 & 2 & 101.33 & 1.53 & 1.45\\
10  & 90.52 & 1 & 101.78 & 1.60 & 1.53\\
12  & 90.00 & 0 & 101.67 & 1.65 & 1.62\\
14  & 89.29 & 4 & 103.00 & 1.70 & 1.75\\
16  & 89.42 & 5 & 101.33 & 1.74 & 1.76\\
20  & 88.61 & 3 & 100.89 & 1.82 & 1.94\\
24  & 88.26 & 2 & 101.33 & 1.89 & 2.08\\
28  & 87.57 & 2 & 99.22 & 1.96 & 2.24\\
32  & 87.01 & 1 & 100.00 & 1.99 & 2.36\\
40  & 86.23 & 4 & 98.78 & 2.09 & 2.64\\
50  & 84.93 & 1 & 98.78 & 2.17 & 2.88\\
64  & 84.17 & 2 & 101.22 & 2.28 & 3.19\\
128 & 81.34 & 3 & 107.44 & 2.57 & 4.16\\
1024 & 73.11 & 23 & 222.01 & 2.83 & 7.55
	\end{tabular}
	\label{tab:no-best-awr}
\end{table}

%\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Variable AWR for Vampire}

In this section we motivate and define a variable clause selection approach.

@@

%%%%%%%
\subsection{Varying AWR (in Vampire)} \label{sec:varying:implementation}

We describe how we vary AWR dynamically and how this is achieved in Vampire. 
%
%If dynamically varying AWR is sometimes useful for finding proofs more quickly, this could be useful as a Vampire option.
In general we would like to describe any possible sequence that the AWR could follow during proof search.
However, some details constrain the design space:
\begin{enumerate}
	\item Changing the AWR too frequently or sharply has little effect, due to the ``balancing'' algorithm --- see Section \ref{sec:introduction}.
	\item A general (configurable) \emph{shape} is more likely to be widely applicable than a specific series of data points.
	\item The shape must extend naturally to an indefinitely-long proof search.
\end{enumerate}

In this work we select two general trends to explore: a trend away from the original (fixed) AWR toward 1:1 (``decay''), and a trend from 1:1 toward the original setting (``converge''). In our investigations we found that even fluctuating sequences had a general trend and these two fixed trends are reasonable approximations of this. We have chosen to use an `original' AWR value to move away from/towards as this original AWR has already been shown to be `useful' in some sense.

@@

Our approach here was restricted by the balancing algorithm used internally, as AWR steps must be discrete and do not take effect immediately.
An alternative approach might use an age/weight probability, rather than a ratio, from which age or weight decisions would be pseudo-randomly (but reproducibly) taken with the use of a seeded PRNG.
This would permit use of continuous age/weight functions, but would also introduce the possibility of ``getting unlucky'' in which a clause preferred by the age/weight probability is not selected due to an improbable-but-possible series of PRNG samples.

>>>>

\documentclass{llncs}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{romannum}

@@

\begin{document}
\maketitle
\begin{abstract}
Modern saturation theorem provers are based on the given-clause algorithm, which iteratively selects new clauses to process. This clause selection has a large impact on the performance of proof search and has been the subject of much folklore.  
The standard approach is to alternate between selecting the \emph{oldest} clause and the \emph{lightest} clause with a fixed, but configurable \emph{age/weight ratio} (AWR).
An optimal fixed value of this ratio is shown to produce proofs significantly more quickly on a given problem, and further that varying AWR during proof search can improve upon a fixed ratio.
Several new modes for the Vampire prover which vary AWR according to a ``shape'' during proof search are developed based on these observations.
The modes solve a number of new problems in the TPTP benchmark set.
\end{abstract}

@@

Currently, the most successful theorem provers (such as Vampire~\cite{vampire}, E~\cite{E}, and SPASS~\cite{Spass}) for first-order logic are saturation-based, utilising the well-known \emph{given-clause algorithm}.
Simply, this algorithm \emph{saturates} a set of clauses by iteratively selecting a clause and performing all non-redundant inferences with it until all clauses have been selected or the empty clause (witnessing inconsistency) has been found.
Clearly, the order in which clauses are selected is key to the performance of the algorithm.
Over the past few decades a certain amount of folklore has built up around the best methods for clause selection and recent work by \sandm{} \cite{clause-selection-heuristics} systematically studied these.
Our work extends this study with new results and also introduces the concept of a \emph{variable} clause selection strategy (one that changes over time), instantiated with two simple patterns (or \emph{shapes}) that prove to be pragmatically useful.

Clause selection strategies that alternate between selecting clauses based on age (i.e. in a first-in-first-out manner) and weight (i.e. those with the fewest symbols first) are the subject of this work.
It was confirmed by \sandm{} that alternating these two heuristics outperforms either by itself.
The ratio of these selections is the \emph{age/weight ratio} (AWR), as this is the terminology employed by the Vampire theorem prover, the vehicle for our study.

After covering relevant background material in Section~\ref{sec:background} the remainder of the paper makes two main contributions.
Firstly, Section~\ref{sec:awr:study} experimentally confirms the folklore that (\romannum{1}) the choice of age/weight ratio often has a significant effect on the performance of proof search, and (\romannum{2}) there is no ``best'' age/weight ratio: indeed, a large range of pragmatically useful ratios exist.
Section~\ref{sec:varying:study} demonstrates that varying the age-weight ratio over time can achieve better performance than a fixed ratio, and therefore motivates the addition of so-called \emph{age/weight shapes} for varying the ratio over time.
Experiments (Section~\ref{sec:experiments}) with these new options implemented in the Vampire theorem prover show a significant improvement in coverage, proving many new problems unsolvable by any previous configuration of Vampire.

@@

\paragraph{First-Order Logic and Weight.}
Our setting is the standard first-order predicate logic with equality. A formal definition of this logic is not required for this paper but an important notion is that of the \emph{weight} of a clause. In first-order logic,  terms are built from function symbols and variables, literals are built from terms, and clauses are disjunctions of literals.
The weight of a term/literal is the number of symbols (function, variable, or predicate) occurring in it.
The weight of a clause is the sum of the weights of its literals.

@@

\paragraph{The Given Clause Algorithm and AWR Clause Selection.}
To achieve saturation, the \emph{given clause algorithm} organises the set of clauses into two sets: the \Active\ clauses are those that have been active in inferences, and the \Passive\ clauses are those that have not. Typically, a further \Unprocessed\ set is required in order to manage the clauses produced during a single iteration of the loop. 
Realisations of the given clause algorithm generally differ in how they organise simplifications. There are two main approaches (both implemented by Vampire, originally found in the eponymous theorem provers Otter~\cite{otter} and \textsc{Discount}~\cite{discount}): the Otter loop uses both \Active\ and \Passive\ for simplifications, whereas the Discount loop uses only \Active.

The algorithm is centred around the \emph{clause selection} process. As previously mentioned, there are two main heuristics for this:
\begin{itemize}
	\item \emph{By Age (or First-in/First-out)} clause selection prefers the oldest clause (produced earlier in proof search), simulating a \emph{breadth-first} search of the clause space. In Vampire the age of a clause is the number of inferences performed to produce it (input clauses have age 0).
	\item \emph{By Weight (or symbol-counting)} clause selection prefers the lightest clause. The intuition behind this approach is that the sought empty clause has zero symbols and lighter clauses are in some sense closer to this. Furthermore, lighter clauses are more general in terms of subsumption and tend to have fewer children, making them less explosive in terms of proof search.
\end{itemize}

\begin{figure}
\begin{tabbing}
~~~~~~\=~~~~\=~~~~\=~~~\=~~~~\=~~~~\=~~~~\=~~~~\=~~~~\=\kill
\>\INPUT: $\Init$: set of clauses\semicol, $a:w$ age-weight ratio \\
\>\VAR\ $\Active$, $\Passive$, $\Unprocessed$: set of clauses\semicol \\
\>\VAR\ $\Given$, $\New$: clause\semicol\\
\>$\Active\ \assign\ \emptyset$\semicol \\
\>$\Unprocessed\ \assign\ \Init$\semicol \\
\Mark
\>$\Balance\ \assign\ 0$\semicol\\
\>\LOOP\\ 
\>\>\WHILE\ $\Unprocessed \neq \emptyset$ \\
\>\>\>$\New \assign \Pop(\Unprocessed)$\semicol \\ 
\>\>\>\IF\ $\New = \emptyclause$ \THEN\ \RETURN\ \textit{unsatisfiable}\semicol \\
\>\>\>\IF\ $\Retained(\New)$ \THEN \` (* retention test *) \\ %\inc
\>\>\>\>simplify $\New$ by clauses in \Active \semicol 
                        \` (* forward simplification *) \\
\>\>\>\>\IF\ $\New = \emptyclause$ \THEN\ \RETURN\ \textit{unsatisfiable}\semicol \\
\>\>\>\>\IF\ $\Retained(\New)$ \THEN
                        \` (* another retention test *) \\
\>\>\>\>\>simplify $\Active$ using $\New$ \semicol
                        \` (* backward simplification *) \\
\>\>\>\>\>move the simplified clauses to $\Unprocessed$\semicol \\
\>\>\>\>\>add $\New$ to $\Passive$\\ %\dec 
\>\>\IF\ $\Passive = \emptyset$ \THEN\
          \RETURN\ \textit{satisfiable} or \textit{unknown}\\
\Mark
\>\>\IF\ $\Balance > 0$ \THEN\\
\Mark
\>\>\>\>$\Given\ \assign\ $ lightest clause in \Passive\semicol\\
\Mark
\>\>\>\>$\Balance \assign\ \Balance - w$\semicol\\
\Mark
\>\>\ELSE\\
\Mark
\>\>\>\>$\Given\ \assign\ $ oldest clause in \Passive\semicol\\
\Mark
\>\>\>\>$\Balance \assign\ \Balance + a$\semicol\\
\>\>move $\Given$ from $\Passive$ to $\Active$\semicol \\
\>\>$\Unprocessed \assign \Infer(\Given,\Active)$\semicol \` (* generating inferences *)\\
\end{tabbing}

@@

\sandm~show that alternating these heuristics is beneficial. In Vampire this alternation is achieved by an age/weight ratio (AWR) implemented by a simple \emph{balancing} algorithm. The balance is initialised to 0 and used as follows: a negative balance means that a clause should be selected by age, whereas a positive balance means that a clause should be selected by weight; given a ratio of $a:w$ the balance is incremented by $a$ when selecting by age and decremented by $w$ when selecting by weight.
Figure~\ref{fig:discount} gives the Discount algorithm along with balance-based AWR clause selection. The lines relevant to clause selection are marked with \Mark.

@@

Vampire's portfolio mode also includes an additional option relevant to clause selection: the \texttt{--nongoal\_weight\_coefficient} option specifies a multiplier to apply to the weight of non-goal clauses, thus preferring clauses in or derived from the problem conjecture in clause selection.
Use of this heuristic is orthogonal to the age/weight ratio and is not investigated further here.

%%%%%%
\paragraph{Related Work.}
Many clause selection approaches are taken by other solvers.
Otter 3.3~\cite{otter3} selects \emph{either} by age, by weight or manually.
Prover9~\cite{prover9} allows a configurable age/weight ratio.
E~\cite{E} allows the user to specify an arbitrary number of priority queues and a weighted round-robin scheme that determines how many clauses are picked from each queue.
The default is to use a combination of age and weight selection, although there is also a complex strategy developed by a genetic algorithm~\cite{genetic-heuristics}. 
SPASS~\cite{Spass} uses symbol-counting based clause selection. 
iProver~\cite{Iprover} follows E in having a number of configurable queues but relies mainly on age and weight heuristics in those queues. 
The general idea in this paper of a \emph{varying age/weight ratio over time} is applicable to any ratio-based clause selection strategy, and our specific results apply to those that take a ratio between age and weight.

\section{Optimising Age/Weight Ratios}
\label{sec:awr:study}
Two assumptions from folklore are confirmed experimentally:
\begin{enumerate}
	\item The choice of age/weight ratio often has a significant effect on the performance of proof search.
	\item There is in general no single best age/weight ratio for a given set of problems.
\end{enumerate}
These are supported by the work of \sandm{} but are explored in more depth here.

\subsection{Logarithmic AWR}
Visualising AWR values is more easily achieved if they have a continuous scale.
AWR values are mathematically \(\mathbb{Q}^{+}\), the positive rational numbers, but in practice are more easily visualised logarithmically.
Therefore, the \emph{logarithmic} AWR \(L\) is defined in terms of age \(A\) and weight \(W\) as
\[
	L = \log_2{\left(\frac{A}{W}\right)}
\]
As \(L\) tends to positive infinity, Vampire selects only by weight, whereas if \(L\) tends to negative infinity Vampire selects only by age.
\(L = 0\) represents the middle ground of a 1:1 age/weight ratio.
Note that the balancing algorithm used by Vampire does not make use of this value (it still requires two numbers) but the quantity is used in this work to show continuous AWR values.

\begin{figure}[t]
	\centering
	\includegraphics[width=0.8\textwidth]{example-optimal-awr}
	\caption{
		The number of given-clause loops reported by Vampire after finding a proof with 1-second runs on a TPTP problem, \texttt{PRO017+2}.
In between the peaks on either side, the function of \(L\) is discontinuous with large peaks and troughs, but follows an approximate trend and settles toward the global optimum.
\texttt{PRO017+2} exhibits typifying behaviour for TPTP, based on manual inspection of several hundred such plots.
}
	\label{fig:example-optimal-awr}
\end{figure}

\subsection{Experiments}
As an initial illustrative example of how varying the AWR effects the number of clauses required to be processed before a proof is found consider Figure \ref{fig:example-optimal-awr}. This demonstrates the effect that varying AWR can have: a smaller number of activations means that fewer clauses were processed, which in general means that a proof was found faster\footnote{
It should be noted that if a small number of clauses are extremely expensive to process it may be slower than a larger number of less-expensive clauses, but in general this is a good heuristic measure for prover performance.
It also avoids reproducibility issues involved with using system timing approaches.
}.
On the problem shown, a good AWR value is over 400\% better by this metric than the worst AWR value.

\begin{table}[t]
	\caption{
Relative performance gain, showing the ratio in activations between the best AWR setting for a given problem and another base setting.
A comparison is drawn between 1:1 (Vampire's default), 1:5 (the best-behaved from \sandm), and the worst setting for the problem.
Where the problem is not solved at all by the base setting, it is ignored.
}
	\centering
	\begin{tabular}{c | c | c c}
		Base Setting & \% Maximum Gain & \% Mean Gain & (Standard Deviation)\\
		\hline
		1:1 & 13,356 & 126 & 163 \\
		1:5 & 13,367 & 144 & 170 \\
		(worst) & 22,201 & 395 & 760
	\end{tabular}
	\label{tab:awr-improvement}
\end{table}

This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not currently support (e.g. higher-order problems).
Vampire ran for 1 second in default mode with the \texttt{discount} saturation algorithm\footnote{The default LRS~\cite{LRS} saturation algorithm can be non-deterministic.} using a sensible set of AWR values (see Table \ref{tab:no-best-awr}) --- these are the values used in Vampire's portfolio mode.
These tend to favour weight-first over age-first as this has been experimentally shown to be preferable.
Problems not solved by any of these, or those solved trivially (e.g in preprocessing) are removed.
The whole set yielded data for 7,947 problems.

The first result is that choosing a good AWR value for a problem is well-rewarded.
Table~\ref{tab:awr-improvement} summarises the impact that choosing the best AWR can have.
Compared to the default, Vampire can perform, on average, 1.26 times fewer activations, which is modest but (as Table \ref{tab:no-best-awr} shows) just under 10\% of problems are no longer proven by choosing the default.
It is more relevant to note that there are cases where Vampire can do \emph{much} better by selecting a different AWR value.
Therefore, choosing a better AWR value \emph{can} go from no solution to a solution and \emph{can} do so faster, but not necessarily.
In the worst case (choosing the pessimal AWR value) Vampire performs almost 4 times as many activations.

The second result is that there is no ``best'' AWR across this full set of problems.
\emph{Drop in performance} is defined to be how many times more activations were required for a proof under a given AWR, compared to the best AWR.
Table \ref{tab:no-best-awr} shows, for each AWR value, the \% of problems solved, the number solved uniquely, and the maximum and mean drop in performance.
No AWR value solves all problems, with the best being 1:5. 
A ratio of 1:4 produces an unusually small maximum performance drop.
\sandm~found that 1:5 had a similar property, but this might be explained by differences in prover and test environment. It is interesting to note that the extreme AWR values solve fewer problems overall but solve the most uniquely. This is typical in saturation-based proof search: approaches that do not perform well in general may perform well in specific cases where the general approach does not.

@@

\begin{table}[t]
	\caption{
Per-AWR value results on 1 second runs over 7,947 TPTP problems.
	}
	\centering
	\begin{tabular}{l | c | c | c | c  c }
		AWR & \% Solved & Uniques & \% Maximum Drop & \% Mean Drop & (Standard Deviation)\\
\hline
8:1 & 85.25 & 16 & 15,067 & 137 & 198\\
5:1 & 86.10 & 1 & 12,222 & 133 & 164\\
4:1 & 86.93 & 1 & 10,144 & 132 & 142\\
3:1 & 87.63 & 2 & 10,500 & 129 & 141\\
2:1 & 88.62 & 3 & 11,267 & 127 & 145\\
3:2 & 89.83 & 2 & 11,989 & 127 & 151\\
5:4 & 89.98 & 4 & 12,500 & 126 & 155\\
1:1 & 90.56 & 4 & 13,356 & 126 & 163\\
2:3 & 91.20 & 9 & 14,767 & 128 & 179\\
1:2 & 91.68 & 0 & 16,267 & 131 & 197\\
1:3 & 91.81 & 5 & 19,056 & 137 & 230\\
1:4 & 91.85 & 3 & \textbf{1,741} & 138 & 67\\
1:5 & 92.00 & 2 & 13,367 & 144 & 170\\
1:6 & 91.57 & 1 & 10,644 & 147 & 146\\
1:7 & 91.49 & 1 & 10,489 & 149 & 144\\
1:8 & 91.09 & 2 & 10,133 & 153 & 145\\
1:10& 90.52 & 1 & 10,178 & 160 & 153\\
1:12& 90.00 & 0 & 10,167 & 165 & 162\\
1:14& 89.29 & 4 & 10,300 & 170 & 175\\
1:16& 89.42 & 5 & 10,133 & 174 & 176\\
1:20& 88.61 & 3 & 10,089 & 182 & 194\\
1:24& 88.26 & 2 & 10,133 & 189 & 208\\
1:28& 87.57 & 2 & 9,922 & 196 & 224\\
1:32& 87.01 & 1 & 10,000 & 199 & 236\\
1:40& 86.23 & 4 & 9,878 & 209 & 264\\
1:50& 84.93 & 1 & 9,878 & 217 & 288\\
1:64& 84.17 & 2 & 10,122 & 228 & 319\\
1:128&81.34 & 3 & 10,744 & 257 & 416\\
1:1024&73.11 & 23 & 22,201 & 283 & 755
	\end{tabular}
	\label{tab:no-best-awr}
\end{table}

%\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Variable AWR for Vampire} 
This section motivates and defines a clause selection approach which varies the AWR value over time.

@@

%%%%%%%
\subsection{Varying AWR (in Vampire)} \label{sec:varying:implementation}
An implementation of dynamically-varying AWR values in Vampire is described below.
In general any possible sequence that the AWR could follow during proof search can be used.
However, some details constrain the design space:
\begin{enumerate}
	\item Changing the AWR too frequently or sharply has little effect, due to the ``balancing'' algorithm --- see Section \ref{sec:introduction}.
	\item A general (configurable) \emph{shape} is more likely to be widely applicable than a specific series of data points.
	\item The shape must extend naturally to an indefinitely-long proof search.
\end{enumerate}

In this work two general trends are explored: a trend away from a given \emph{start} AWR toward 1:1 (``decay''), and a trend from 1:1 toward a given \emph{end} AWR (``converge'').
Investigation showed that even fluctuating sequences had a general trend, and further that these two fixed trends are reasonable approximations of these trends.
The start/end AWR values are taken from the portfolio mode: these values are known to be useful in a fixed-AWR context, and while this may not generalise to a dynamic-AWR context, it is a useful starting point pending integration of AWR shape parameters into strategy scheduling.

@@

Our approach here was restricted by the balancing algorithm used internally, as AWR steps must be discrete and do not take effect immediately.
An alternative approach might be to use an age/weight probability, rather than a ratio, from which age or weight decisions would be pseudo-randomly (but reproducibly) taken with the use of a seeded pseudo-random number generator, permitting use of continuous age/weight functions.
